Tackling irreproducible enrichment analysis

### 2024-03-12 manuscript/manuscript.Rmd
ms work

@@

Irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only four articles exhibited a high degree of concordance, while seven exhibited major
discrepancies, which we mainly ascribe to deficiencies in methodological reporting and lack of
preservation of past DAVID versions.
Better reporting and a long-term view of reproducibility are needed, which are satisfied with
scripted workflows and shared code.
We provide a step-by-step beginner-friendly enrichment analysis protocol
exemplifying current best practices including literate programming, version control, containerisation,
documentation and persistent sharing of data and code.

@@

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID (Database for Annotation,
Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
The basis for these selection criteria is that this combination represents the most common application
of enrichment analysis in our previous study, which could be reproduced without the need to
reanalyse the raw dataset from scratch. These articles are listed in the bibliography
[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].

@@

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring @Zeeberg2004-na.
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID v6.8 as at this time v6.7 was not available
(conducted March-May 2022) [@Sherman2022-hf].

@@

Each article was critically evaluated and given a score of one (1) to three (3).
Articles rated 1 showed overall poor reproducibility.
For example, most GO and/or KEGG terms mentioned in the results and discussion were not confirmed
upon reproduction, and the conclusions of the study were overall not supported.
Articles were rated 2 when at least 50% of the terms are observed as significant after replication
according to the original study's significance threshold, and most conclusions were supported.
Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
shown in the article observed again in reproduction, and all conclusions were supported in reproduction.

@@

As reproducibility is associated with the quality of documentation, we collected key methodological information
about how the enrichment analysis was done and present in Table 1.
From this, we can see that only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a background list was used for enrichment analysis.
We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.

@@

We attempted reproduction according to the methods described in the respective articles, with the limitation
that we had to use DAVID v6.8 as v6.7 was no longer available.
The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
Seven studies were classified as low reproducibility, while 9 were classified as medium and just four
were classified as high.
Justifications for these classifications are provided in Table S1.

@@

Taken together, the ability to reproduce DAVID enrichment analysis using the supplementary gene
lists was highly variable.
As DAVID version 6.8 has been retired, these and approximately 20,000 other PubMed articles have enrichment analyses
that are no longer reproducible with the original tools.
These observations highlight the need for better methodological reporting to reduce ambiguity in methods sections
and suggest that web-based tools without long-term reproducibility features are no longer suitable for inclusion in
scientific publications.

@@

The results show that poor reporting and lack of preservation of web-tools is to blame for poor
reproducibility.
There is a need to bring enrichment analysis in line with bioinformatics best practices, so
we present a beginner-friendly protocol that guides researchers through the process of
literate scripting, version control, containerisation, documentation and code/data sharing.

@@

With ORA, we observed 201 up-regulated and 99 down-regulated gene sets (FDR<0.05).
We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo small expression changes.
This led to the identification of up-regulated pathways related to nuclear RNA import and
glucokinase regulation; while down-regulated pathways
were associated with interferon signaling, interleukin signaling and NADPH oxidase
(host defence) (**Figure 4C**).

@@

The problem of bioinformatic resources becoming unavailable is not a new one, being thoroughly described a
decade ago [@Hennessey2013-tj].
We are worried that a reliance on web tools that lack open source code or containers can leave
researchers in violation of their data management responsibilities, as preservation mandates
include software as well as original research data
[@NHMRC2019-wu ; @Harvard_2023-cl].

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
risk of becoming unavailable without notice due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma and allows users to download the computer script to facilitate
future non-interactive reproducibility @Powell2019-lv.
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the highly reproducible
approach detailed here.

@@

Although we hope that highly reproducible enrichment analysis will become more widespread, we do not
anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
Currently, reward structures in academia favour career progression for individuals who publish many
articles as opposed to individuals who publish fewer but more rigorously conducted works.
Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
badging of articles that comply with best practices @Kidwell2016-pm, and recognition of research rigour
in track record evaluations.

>>>>

Alt: *Tackling irreproducible pathway enrichment analysis*

@@

Irreproducible research is a significant problem that wastes resources and risks
undermining the public perception of science.
Previous work has highlighted that published enrichment analyses frequently suffer from statistical
and reporting flaws.
We sought to determine whether this translates into irreproducibility by examining whether the
findings of 20 open-access articles published in 2019 describing enrichment analysis with the
popular DAVID suite were reproducible.
We find that only four articles exhibited a high degree of concordance, while seven exhibited major
discrepancies.
These inconsistencies can mostly be ascribed to deficiencies in methodological reporting and lack of
preservation of past DAVID versions.
Better reporting and a long-term view of reproducibility are needed.
Scripted workflows and shared code are a potential solution to this problem, although there is a
steep learning curve for novice analysts.
We provide a step-by-step beginner-friendly enrichment analysis protocol
exemplifying current best practices including literate programming, version control, containerisation,
documentation and persistent sharing of data and code.

@@

From a list of 186 PubMed central articles using enrichment analysis that we examined previously
@Wijesooriya2022-li, we selected 20 articles that fulfilled the following criteria; (1) the
organism under study is human; (2) the tool used was DAVID (Database for Annotation,
Visualization, and Integrated Discovery) @Huang2007-yw ; (3) the omics type was gene
expression (array or RNA-seq); and (4) that the gene list was provided in the supplementary data.
The basis for these selection criteria is that this combination represents the most common application
of enrichment analysis according to our previous study, which could be reproduced without the need to
reanalyse raw data. These articles are listed in the bibliography
[@Liu2019-nm; @Zhang2019-bq; @Zhang2019-zs; @Wang2019-il; @Balusamy2019-ue; @Ahn2019-wl; @Chai2019-cn;
@Guo2019-fx; @Alebrahim2019-zc; @Hong2019-ym; @Han2019-ix; @McSweeney2019-ij; @Wang2019-yw; @Pan2019-iv;
@Wang2019-jy; @Weigt2019-mk; @Najafi2019-ce; @Feltrin2019-to; @Wang2019-xx; @Sun2019-xr].

@@

The lists of genes were collected from the supplementary files and stored in a spreadsheet, paying
special attention to preventing gene name errors from occurring [@Zeeberg2004-na,Ziemann2016-pv, Abeysooriya2021-uz].
The reproducibility analysis was performed using the respective methodologies stated in the methods
sections of the 20 articles using DAVID v6.8, as at this time v6.7 was not available
(conducted March-May 2022) [@Sherman2022-hf].

@@

Each article was critically evaluated and given a score of one (1) to three (3).
Articles rated 1 showed overall poor reproducibility.
For example, less than 50% of GO and/or KEGG terms mentioned in the results and discussion were confirmed
upon reproduction, and the conclusions of the study were overall not supported.
Articles were rated 2 when at least 50% of the terms are observed as significant after replication
according to the original study's significance threshold, and most conclusions were supported.
Articles were rated 3 when they showed a high level of reproducibility, with 80% or more of the terms
shown in the article observed again in reproduction, and all conclusions were supported in reproduction.

@@

As reproducibility is closely associated with the quality of documentation, we collected key methodological
information about how the enrichment analysis was done and present in Table 1.
From this, we can see that only seven articles reported the version of DAVID that was used.
Version 6.7 was reportedly used in two articles while version 6.8 was specified in five.
Correction of p-values for multiple testing was specified or inferred in 11 articles, while the results of
the remaining nine articles did not take into consideration the importance of p-value correction for
minimising false positives.
Of the 20 articles, none mentioned explicitly that a custom background list was used for enrichment analysis.
We also examined whether the gene selection criteria were unambiguously clear, which was true in 16 articles.

@@

We attempted reproduction according to the methods described in the respective articles, with the limitation
that we had to use DAVID v6.8 as v6.7 was no longer available.
The 20 articles were classified based on the degree of reproducibility (**Figure 1**).
Seven studies were classified as low reproducibility, while nine were classified as medium and just four
were classified as high.
Justifications for these classifications are provided in Table S1.

@@

The ability to reproduce DAVID enrichment analysis using the supplementary gene
lists was highly variable.
The version of DAVID used is likely to be a big contributor to irreproducibility as we know that there are major
differences in the underlying annotation set (knowledgebase) used by version 6.7 and 6.8.
This is supported by the fact that none of the v6.7 studies received a high reproduicibility score.
On the other hand these versions did not receive regular updates, so the knowledgebases were static for relatively
long times; v6.7 (Jan. 2010 - Oct 2020) and v6.8 (Oct 2016 - Jun 2022).
We can then deduce that knowledgebase version is likely not the cause of irreproducibility for PMC6561911,
PMC6539328 and PMC6425008; studies that reportedly used v6.8 but received low reproducibility scores.

We note that DAVID version 6.8 was retired in June 2022, so these 20 studies are currently no longer reproducible
with the original tools.
The implications of this are outlined in the Discussion section below.

@@

Our results emphasise the need for better methodological reporting to reduce ambiguity in methods sections
and highlight that webtools can suddenly become unavailable.
Although there are various solutions to these issues (refer to the Discussion section), here we will outline
a potential solution whose design is guided by current best practices
[@Tan2010-is; @Sandve2013-lk; @Piccolo2016-ji; @Gruning2018-sg; @Ziemann2023-dg].
Our aim is to provide a beginner-friendly guide to enrichment analysis that will be
reproducible for many years to come.

@@

With ORA, we observed 201 up-regulated and 99 down-regulated gene sets (FDR<0.05).
We prioritised enrichment results by enrichment score, a surrogate measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo small expression changes.
This led to the identification of up-regulated pathways related to nuclear RNA import and
glucokinase regulation; while down-regulated pathways
were associated with interferon signaling, interleukin signaling and NADPH oxidase
(host defence) (**Figure 4C**).

@@

The problem of bioinformatic resources including web services becoming unavailable is not a new one, being
raised over a decade ago [@Hennessey2013-tj] and revisited in 2020 [@Kern2020-iz].
We are worried that a reliance on web tools that lack open source code or containers can leave
researchers in violation of their data management responsibilities, as preservation mandates
include software as well as original research data
[@NHMRC2019-wu ; @Harvard_2023-cl].

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
risk of becoming unavailable without notice due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma and allows users to download the computer script to facilitate
future non-interactive reproducibility [@Powell2019-lv].
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions [@Ge2020-qe], facilitating preservation and aiding future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the highly reproducible
approach detailed here.

@@

Although we hope that highly reproducible enrichment analysis will become more widespread, we do not
anticipate any dramatic changes in researcher behaviour unless the motivating factors are addressed.
Currently, reward structures in academia favour career progression for individuals who publish many
articles as opposed to individuals who publish fewer but more rigorously conducted works.
Solutions to this could include mandates by funding agencies, institutions or journals [@Garijo2013-zb],
badging of articles that comply with best practices [@Kidwell2016-pm], and recognition of research rigour
in track record evaluations.
